Copulas and their potential for ecology

### 2019-09-19 Paper.Rmd
done a roudn of edits up through Concepts and Methods for Q1

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables; some of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. C and D show normalized ranks of A, B, 
respectively. (E,F) Reflections
of C, D, respectively, about a vertical axis. Subscripts h and v 
in the axis labels stand for "horizontal" and "vertical".\label{fig_cop_pedag1}}
\end{center}
\end{figure}
<!--***DAN: When I cut panels out of this figure (saving them for use in a 
later figure), I also cut the caption, and here is the cut stuff for later use:
(C, D) Contours 
of probability density functions (pdf) of Clayton (C) and survival-Clayton 
(D) copulas with parameter 5 (see sections on methods for Q1), together
with samples from the corresponding distributions. (E,F) Reflections
of C, D, respectively, about a vertical axis. (G) Contours of the
pdf of a normal copula (methods for Q1) with 
parameter chosen to
have the same Spearman correlation as A-D, together with a sample.
(H) The portion of the Spearman correlation within a band 
can be computed
for any band (methods for Q1) to describe how dependence
varies from one part of the two distributions to another. Diagonal lines
show two bands, the data in the left/lower band showing stronger
dependence than the right/upper band.
Subscripts h and v 
stand for "horizontal" and "vertical".
-->

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*; copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the section on Background below. Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cummulative distribution function (cdf) of the $x_i$ and $y_i$,
respectively. See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described in the section on Background below, 
after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals. Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables; 
they provide information solely on the dependence between those variables.

@@

The goal of this paper is to explore the potential for 
applications of copulas in ecology. We provide a basic introduction
to copula concepts in the Background section below, but this is a sketch
only - readers can find abbreviated [@Anderson2018] 
and comprehensive introductions elsewhere
[@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. Instead our goal is to 
help determine to what extent ecological understanding may benefit generally from
using copulas. Copulas have already been used very effectively 
in a few studies to illuminate ecological phenomena 
[@Valpine2014; @Anderson2018; @Popovic2018], but such usage is
still rare. Our results suggest that environmental, 
ecological and evolutionary processes 
may commonly generate and transmit complex dependence structures, including tail 
dependence, and that greater use of copulas can illuminate the underlying processes.

@@

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
associations distinct from those of a normal distribution, and in particular
do they show asymmetric tail associations? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from that standard or default cases. 
(Q2) If the answer to
Q1 is "yes," then what are some possible causes/mechanisms of non-normal
copula structure and asymmetric tail associations in ecology? And (Q3) what are the 
consequences of non-normal copula
structure and asymmetric tail associations for ecological understanding and
applications? After a Background section (the second section), we address Q1 
(third through fifth sections of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sixth and seventh sections of the paper), using 
simple models. We address Q3 (eighth and ninth sections of the paper) using 
both data and models. The final section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have been used very effectively, but rarely so far in ecology [@Valpine2014; 
@Anderson2018; @Popovic2018]. 
We argue that benefits of wider usage of complete 
copula descriptions of dependence in ecology will be substantial.

@@

We give a brief introduction to copulas, focusing, for simplicity, on bivariate copulas
and on concepts needed
for the rest of the paper. See @nelsen2006_copula, @joe2014_dependence, and
@MaiScherer2017 for general background, and see also @Anderson2018, part of 
which is a short introduction to 
copulas for ecologists. A bivariate *copula* can be defined as a bivariate distribution
function with both margins uniform on $(0,1)$ (@joe2014_dependence, p. 7). 
This will be the
distribution function of a bivariate random variable with uniform marginals, and 
the terminology *copula* is sometimes also applied to the random variable. 
A foundational theorem of
@sklar1959_theorem (see also @MaiScherer2017, p. 16) 
says that if $F$ is the bivariate distribution function
of a random vector $(X,Y)$, with margins $F_X$ and $F_Y$, then there 
exists a copula $C$
such that for all $(x,y)$ in the Euclidean plane, $F(x,y)=C(F_X(x),F_Y(y))$.
The theorem also says $C$ is unique if $F_X$ and $F_Y$ are continuous.
Thus $C$ couples the bivariate distribution, $F$, with the two marginal distributions. 
Finally, Sklar's theorem says that if $D$ is any bivariate copula and 
$G_A$ and $G_B$ are univariate distribution functions, then 
$D(G_A(x),G_B(y))$ is a distribution function.
The applications of this study fall within the case in which univariate marginal 
distributions have continuous, strictly 
monotonic distribution functions, and this case is simpler, so we 
henceforth make such an assumption. 
Then, "the" copula (now guaranteed unique) 
associated with $(X,Y)$ is the distribution
function of the random variable $(F_X(X),F_Y(Y))$ 
(Appendix \ref{SM-ContStrictMonoton}). And 
$D(G_A(x),G_B(y))$ is the distribution function of the random variable 
$(G_A^{-1}(U),G_B^{-1}(V))$, where $(U,V)$ is the random
vector with distribution function $D$ and
$G_A^{-1}$ and $G_B^{-1}$ are the inverses of $G_A$ and $G_B$
(Appendix \ref{SM-ContStrictMonoton}). Application of the 
distributions functions of the component random variables translates
between the uniform-marginal context, which isolates the 
dependence information, and the context of the original random variable.

@@

<!--***DAN: I still feel like we have to make this more intuitive for people
at this point. Maybe a figure with two panels showing two copulas on
the left, and two panels shwoing two sets of marginals at the top,
and then four panels showing what you get when each of these combine?
Good idea. Steps: 
DONE split the current Fig 1 so the new Fig 1 has the first 6 panels,
but C-D should only have the points. 
DONE Then make a Fig 2 as above, 
DONE and edit the text in the above para to use that figure. 
STILL HAVE TO DO The rest of the panels of the current Fig 1 will make a new Fig 3 later.
-->

@@

We introduce some standard copula families, as examples of copulas and also
because we will fit these families to data later. The bivariate normal
copula family (henceforth abbreviated N), which consists of copulas of bivariate normal
distributions, is a one-parameter family. The parameter, which we here call $p$,
corresponds to the correlation
of the related bivariate normal distribution, and controls the degree of dependence
between the variables. An N copula with $p=0.7$ was already introduced (Fig.
\ref{PedagogFigCopThry}A). The formula for the
copula is $\Phi_{2,p}(\Phi_{1}^{-1}(u),\Phi_{1}^{-1}(v))$, where $\Phi_1^{-1}$ is the inverse of the
cdf of a (univariate) standard normal and $\Phi_{2,p}$ is the joint cdf of a
bivariate normal distribution with mean $(0,0)$ and covariance matrix having $1$s on the diagonal
and $p$ in the off diagonal positions. However, formulas for copulas are often not as 
directly informative about copula properties as pdf diagrams; Fig. \ref{figped_N} has 
example pdfs correspondng to bivariate N copulas for various values of $p$.
Note that the pdfs are symmetric
around the diagonal line $v=-u+1$ (Fig. \ref{figped_N}), so normal copulas have symmetric 
associations between the two variables in the left and right tails (see
below for a more precise defintion of tail associations). 
The Clayton copula family (henceforth abbreviated C),
of which we also already pictured an example (Fig.
\ref{PedagogFigCopThry}B), is another one-parameter family. In contrast to N copulas, C copulas
have stronger left- than right-tail association. The formula is 
$\left[  \reumax(u^{-p}+v^{-p}-1,0)\right]^{-1/p}$, for parameter $p$; though this again provides 
less direct intuition than do example pdfs (Fig. \ref{figped_C}). The survival
Clayton family (here abbreviated SC) is the symmetric opposite of the C family, showing
stronger right- than left tail association (Fig. \ref{SM-figped_C}). The BB1 copula 
family is a two-parameter family, thereby providing more flexibility: for some parameters
BB1 copulas have stronger left- than right-tail association, and for others the reverse
(Fig. \ref{figped_BB1}). See @nelsen2006_copula and @joe2014_dependence for further
information on these and other copula families.

@@

Having introduced tail association conceptually and referred to it
in examples, we now need a more precise definition of a measure of tail association.
We use the term *tail association* to describe the general idea of the strength of
associations between two variables in the tails of their distributions. One of the
ways this is measured is called *tail dependence* (@joe2014_dependence, section 2.13; 
@nelsen2006_copula, section 5.4), which we here define. Given a random vector $(X,Y)$
with margins $F_X$, $F_Y$ and defining $U=F_X(X)$, $V=F_Y(Y)$, the upper-tail
dependence of $X$ and $Y$ is defined as 
$\lambda_U = \lim_{u \rightarrow 1^-} P[Y>F_Y^{-1}(u) | X>F_X^{-1}(u)]$. This equals
$\lim_{u \rightarrow 1^-} P[U>u | V>u]$, which in turn equals 
$\lim_{u \rightarrow 1^-} P[U>u, V>u]/P[V>u]=\lim_{u \rightarrow 1^-} P[U>u, V>u]/(1-u)$,
which shows upper-tail dependence is defined symmetrically in the two random variables.
Lower tail dependence is defined analogously as 
$\lambda_L = \lim_{u \rightarrow 1^-} P[Y \leq F_Y^{-1}(u) | X \leq F_X^{-1}(u)]$.
Tail dependence is a property of the random variable $(U,V)$, so depends only
on the copula and not on the marginals combined with it. It can be shown 
(@nelsen2006_copula, section 5.4) that $\lambda_L=\lambda_U=0$ for the N copula,
$\lambda_L=2^{-1/p}$ and $\lambda_U=0$ for a C copula with parameter $p>0$,
and $\lambda_L=0$ and $\lambda_U=2^{-1/p}$ for an SC copula with parameter $p>0$.
In the section on methods for Q1 we will introduce
ways tail dependence can be applied to data, by first fitting copulas to the data. We will also 
introduce nonparametric measures of tail association that apply directly to data.

Plots using normalized ranks were proposed in the Introduction as being conceptually
similar to copulas. We here explain the connection. As stated above, the copula associated
with a random variable $(X,Y)$ is the joint distribution function of the random variable
$(F_X(X),F_Y(Y))$. Given a sample $(x_i,y_i)$, $i=1,\ldots,n$, from $(X,Y)$, let $\hat{F}_X$ and
$\hat{F}_Y$ be the empirical cdfs associated with the $x_i$ and $y_i$, respectively.
These are step functions of $x$ and $y$, respectively, that start at $0$ for low values of 
$x$ and $y$ and jump by $1/n$
at each of the data points; for large $n$ they approximate the cdfs $F_X$ and $F_Y$. 
But $\hat{F}_X(x_i)$ equals the normalized rank of $x_i$ times
$n/(n+1)$, and likewise $\hat{F}_Y(y_i)$ equals the normalized rank of $y_i$ times
$n/(n+1)$. So the normalized rank in turn approximates the empirical cdf and therefore 
the cdf. Thus the normalized rank pairs $(u_i,v_i)$ for $i=1,\ldots,n$ can be 
regarded, for large $n$, as a sample from the bivariate pdf associated with the copula of $(X,Y)$,
and the scatterplot of $v_i$ versus $u_i$ can be used to infer aspects of copula structure
(see section on methods for Q1 below).

<!--***DAN:
You are going to have to rearrange the order of some figs in Sup Mat because now the data section comes after Bacground, in which the different copula famlies are introduced
-->

<!--***DAN: put this text in the methods. It won't help people here in the ms to have all
these copulas intro'd. The sole purpose of them is give a variety of options to have a robust model
selection and multi-model inference procedure for tail dependence, so mention them in that context in 
the Methods in a way that make clear we don't need to give a damn about these things:
For our model selection procedure (see below), we used 
several families of copulas exhibiting various 
lower- and upper-tail dependence characteristics 
(Clayton, survival-Clayton, Gumbel, survival-Gumbel, Joe, survival-Joe, Frank,
BB1, survival-BB1,
BB6, survival-BB6, BB7, survival-BB7, BB8, survival-BB8, henceforth
abbreviated C, SC, G, SG, J, SJ, F, BB1, SBB1, BB6, SBB6, BB7, SBB7, BB8, SBB8).
Copulas are implemented in the R packages `copula` and `VineCopula`. 
Visualizations of these families (Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8})
can be compared with normalized rank plots to estimate visually 
which families may effectively model a given bivariate dataset. 
If LT and UT denote 
lower- and upper-tail dependence measures, their difference measures asymmetry of 
tail dependence.
See @nelsen2006_copula, @joe2014_dependence for further background. -->

\begin{figure}[!h]
\begin{center}
\includegraphics[width=\textwidth]{./Results/PedagogSuppMat_Normal.pdf}
\caption[Example N copulas]{Log-transformed pdfs (A-E) and samples (F-J) from example N copulas. 
S is Spearman correlation; p is the value of the parameter for the N 
family (it is a 
one-parameter family); and LT and UT are the measures of lower- and upper-tail dependence, respectively (see Background section for definition).\label{figped_N}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=\textwidth]{./Results/PedagogSuppMat_Clayton.pdf}
\caption[Example C copulas]{Log-transformed pdfs (A-E) and samples (F-J) from example C 
copulas. S is Spearman correlation; p is the value of the parameter for the C family (it 
is a one-parameter family); and LT and UT are the measures of lower- and 
upper-tail dependence, 
respectively (see Background section for definition).\label{figped_C}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=\textwidth]{./Results/PedagogSuppMat_BB1.pdf}
\caption[Example BB1 copulas]{Log-transformed pdfs for example BB1 copulas. K is Kendall correlation; p1 and p2
denote the two parameters of the family (it is a 
two-parameter family); and LT and UT are our measures of lower- and upper-tail dependence, 
respectively (see Background section for definition).\label{figped_BB1}}
\end{center}
\end{figure}

@@

<!--***DANr: I moved some text from here and have not smoothed the transtion,
if it needs smoothing-->

We addressed Q1 principally via a model selection procedure in which the 
copulas listed above were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, we produced normalized ranks 
$u_i$ and $v_i$ as in the Introduction. 
Second, we tested the independence of the $u_i$ and $v_i$ 
(using `BiCopIndTest` in the `VineCopula` package). 
Our model 
selection algorithms were ineffective if data could not be distinguished from independent data 
since many of the copula families we considered include the independent copula. 
If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the 16 bivariate copula families listed previously to the 
normalized ranks via maximum likelihood, using `BiCopEst` in `VineCopula`, 
then obtaining AIC and BIC values and accompanying model weights $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. `BiCopEst` also provided the lower- and 
upper-tail dependence of the best-fitting member of each family. 
Fourth, $\text{AIC}_{\text{w}}$ values were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods give the *relative* support of several models, but do not 
indicate whether any of the models are an objectively good fit. To test that, 
and as our final step, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure (`BiCopGofTest` of `VineCopula`). The procedure performed 
one test based on a Cramer-von Mises statistic and another based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
a run using $100$ bootstraps was performed; if
the $p$-value from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

>>>>

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables; some of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. C and D show normalized ranks of A, B, 
respectively. (E,F) Reflections
of C, D, respectively, about a vertical axis. Subscripts h and v 
in the axis labels stand for "horizontal" and "vertical".\label{fig_cop_pedag1}}
\end{center}
\end{figure}

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*; copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the section on Background below. Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cummulative distribution function (cdf) of the $x_i$ and $y_i$,
respectively. See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described in the section on Background below, 
after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals \cite{Genest2007}. 
Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables; 
they provide information solely on the dependence between those variables.

@@

The goal of this paper is to explore the potential for 
applications of copulas in ecology. We provide a basic introduction
to copula concepts in the Background section below, but this is a sketch
only - readers can find abbreviated [@Anderson2018; @Genest2007] 
and comprehensive introductions elsewhere
[@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. Instead our goal is to 
help determine to what extent ecological understanding may benefit generally from
using copulas. Copulas have already been used very effectively 
in a few studies to illuminate ecological phenomena 
[@Valpine2014; @Anderson2018; @Popovic2018], but such usage is
still rare. Our results suggest that environmental, 
ecological and evolutionary processes 
may commonly generate and transmit complex dependence structures, including tail 
dependence, and that greater use of copulas can illuminate the underlying processes.

@@

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
associations distinct from those of a normal distribution, and in particular
do they show asymmetric tail associations? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from the standard or default case. 
(Q2) If the answer to
Q1 is "yes," then what are some possible causes/mechanisms of non-normal
copula structure and asymmetric tail associations in ecology? And (Q3) what are the 
consequences of non-normal copula
structure and asymmetric tail associations for ecological understanding and
applications? After a Background section (the second section), we address Q1 
(third through fifth sections of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sixth and seventh sections of the paper), using 
simple models. We address Q3 (eighth and ninth sections of the paper) using 
both data and models. The final section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have been used very effectively, but rarely so far in ecology [@Valpine2014; 
@Anderson2018; @Popovic2018]. 
We argue that benefits of wider usage of complete 
copula descriptions of dependence in ecology will be substantial.

@@

We give a brief introduction to copulas, focusing, for simplicity, on bivariate copulas
and on concepts needed
for the rest of the paper. See @nelsen2006_copula, @Genest2007, @joe2014_dependence, and
@MaiScherer2017 for general background, and see also @Anderson2018, part of 
which is a short introduction to 
copulas for ecologists. A bivariate *copula* can be defined as a bivariate distribution
function with both margins uniform on $(0,1)$ (@joe2014_dependence, p. 7). 
This will be the
distribution function of a bivariate random variable with uniform marginals, and 
the terminology *copula* is sometimes also applied to the random variable. 
A foundational theorem of
@sklar1959_theorem (see also @MaiScherer2017, p. 16) 
says that if $F$ is the bivariate distribution function
of a random vector $(X,Y)$, with margins $F_X$ and $F_Y$, then there 
exists a copula $C$
such that for all $(x,y)$ in the Euclidean plane, $F(x,y)=C(F_X(x),F_Y(y))$.
The theorem also says $C$ is unique if $F_X$ and $F_Y$ are continuous.
Thus $C$ couples the bivariate distribution, $F$, with the two marginal distributions. 
Finally, Sklar's theorem says that if $D$ is any bivariate copula and 
$G_A$ and $G_B$ are univariate distribution functions, then 
$D(G_A(x),G_B(y))$ is a distribution function.
The applications of this study fall within the case in which univariate marginal 
distributions have continuous, strictly 
monotonic distribution functions, and this case is simpler, so we 
henceforth make such an assumption. 
Then, "the" copula (now guaranteed unique) 
associated with $(X,Y)$ is the distribution
function of the random variable $(F_X(X),F_Y(Y))$ 
(Appendix \ref{SM-ContStrictMonoton}). And 
$D(G_A(x),G_B(y))$ is the distribution function of the random variable 
$(G_A^{-1}(U),G_B^{-1}(V))$, where $(U,V)$ is the random
vector with distribution function $D$ and
$G_A^{-1}$ and $G_B^{-1}$ are the inverses of $G_A$ and $G_B$
(Appendix \ref{SM-ContStrictMonoton}). Application of the 
distributions functions of the component random variables translates
between the uniform-marginal context, which isolates the 
dependence information, and the context of the original random variable.

@@

@@

We introduce some standard copula families, as examples of copulas and also
because we will fit these families to data later. The bivariate normal
copula family (henceforth abbreviated N), which consists of copulas of bivariate normal
distributions, is a one-parameter family. The parameter, which we here call $p$,
corresponds to the correlation
of the related bivariate normal distribution, and controls the degree of dependence
between the variables. An N copula with $p=0.7$ was already introduced (Fig.
\ref{PedagogFigCopThry}A). The formula for the
copula is $\Phi_{2,p}(\Phi_{1}^{-1}(u),\Phi_{1}^{-1}(v))$, where $\Phi_1^{-1}$ is the inverse of the
cdf of a (univariate) standard normal and $\Phi_{2,p}$ is the joint cdf of a
bivariate normal distribution with mean $(0,0)$ and covariance matrix having $1$s on the diagonal
and $p$ in the off diagonal positions. However, formulas for copulas are often not as 
directly informative about copula properties as pdf diagrams; Fig. \ref{figped_N} has 
example pdfs corresponding to bivariate N copulas for various values of $p$.
Note that the pdfs are symmetric
around the diagonal line $v=-u+1$ (Fig. \ref{figped_N}), so normal copulas have symmetric 
associations between the two variables in the left and right tails (see
below for a more precise defintion of tail associations). 
The Clayton copula family (henceforth abbreviated C),
of which we also already pictured an example (Fig.
\ref{PedagogFigCopThry}B), is another one-parameter family. In contrast to N copulas, C copulas
have stronger left- than right-tail association. The formula is 
$\left[  \reumax(u^{-p}+v^{-p}-1,0)\right]^{-1/p}$, for parameter $p$; though this again provides 
less direct intuition than do example pdfs (Fig. \ref{figped_C}). The survival
Clayton family (here abbreviated SC) is the symmetric opposite of the C family, showing
stronger right- than left tail association (Fig. \ref{SM-figped_SC}). The BB1 copula 
family is a two-parameter family, thereby providing more flexibility: for some parameters
BB1 copulas have stronger left- than right-tail association, and for others the reverse
(Fig. \ref{figped_BB1}). See @nelsen2006_copula and @joe2014_dependence for further
information on these and other copula families.

@@

Having introduced tail association conceptually and referred to it
in examples, we now need a more precise definition of a measure of tail association.
We use the term *tail association* to describe the general idea of the strength of
associations between two variables in the tails of their distributions. One of the
ways this is measured is called *tail dependence* (@joe2014_dependence, section 2.13; 
@nelsen2006_copula, section 5.4), which we here define. Given a random vector $(X,Y)$
with margins $F_X$, $F_Y$ and defining $U=F_X(X)$, $V=F_Y(Y)$, the upper-tail
dependence of $X$ and $Y$ is defined as 
$\lambda_U = \lim_{u \rightarrow 1^-} P[Y>F_Y^{-1}(u) | X>F_X^{-1}(u)]$. This equals
$\lim_{u \rightarrow 1^-} P[U>u | V>u]$, which in turn equals 
$\lim_{u \rightarrow 1^-} P[U>u, V>u]/P[V>u]=\lim_{u \rightarrow 1^-} P[U>u, V>u]/(1-u)$,
which shows upper-tail dependence is defined symmetrically in the two random variables.
Lower tail dependence is defined analogously, as 
$\lambda_L = \lim_{u \rightarrow 1^-} P[Y \leq F_Y^{-1}(u) | X \leq F_X^{-1}(u)]$.
Tail dependence is a property of the random variable $(U,V)$, so depends only
on the copula and not on the marginals combined with it. It can be shown 
(@nelsen2006_copula, section 5.4) that $\lambda_L=\lambda_U=0$ for the N copula,
$\lambda_L=2^{-1/p}$ and $\lambda_U=0$ for a C copula with parameter $p>0$,
and $\lambda_L=0$ and $\lambda_U=2^{-1/p}$ for an SC copula with parameter $p>0$.
In the section on methods for Q1 we will introduce
ways tail dependence can be applied to data, by first fitting copulas to the data. We will also 
introduce nonparametric measures of tail association that apply directly to data.

Plots using normalized ranks were proposed in the Introduction as being conceptually
similar to copulas. We here explain the connection. As stated above, the copula associated
with a random variable $(X,Y)$ is the joint distribution function of the random variable
$(F_X(X),F_Y(Y))$. Given a sample $(x_i,y_i)$, $i=1,\ldots,n$, from $(X,Y)$, let $\hat{F}_X$ and
$\hat{F}_Y$ be the empirical cdfs associated with the $x_i$ and $y_i$, respectively.
These are step functions of $x$ and $y$, respectively, that start at $0$ for low values of 
$x$ and $y$ and jump by $1/n$
at each of the data points; for large $n$ they approximate the cdfs $F_X$ and $F_Y$. 
But $\hat{F}_X(x_i)$ equals the normalized rank of $x_i$ times
$n/(n+1)$, and likewise $\hat{F}_Y(y_i)$ equals the normalized rank of $y_i$ times
$n/(n+1)$. So the normalized rank in turn approximates the empirical cdf and therefore 
the cdf. Thus the normalized rank pairs $(u_i,v_i)$ for $i=1,\ldots,n$ can be 
regarded as approximate samples from the bivariate pdf 
associated with the copula of $(X,Y)$,
and the scatterplot of $v_i$ versus $u_i$ can be used to infer aspects of copula structure
(see section on methods for Q1 below).

\begin{figure}[!h]
\begin{center}
\includegraphics[width=\textwidth]{./Results/PedagogSuppMat_Normal.pdf}
\caption[Example N copulas]{Log-transformed pdfs (A-E) and samples (F-J) from example N copulas. 
S is Spearman correlation; $p$ is the value of the parameter for the N 
family (it is a 
one-parameter family); and LT and UT are the measures of lower- and upper-tail dependence, respectively (see Background section for definition). The 
parameter range for the family is $p \in [-1,1]$, and lower- and upper-tail
dependence are always $0$.\label{figped_N}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=\textwidth]{./Results/PedagogSuppMat_Clayton.pdf}
\caption[Example C copulas]{Log-transformed pdfs (A-E) and samples (F-J) from example C 
copulas. S is Spearman correlation; $p$ is the value of the parameter for the C family (it 
is a one-parameter family); and LT and UT are the measures of lower- and 
upper-tail dependence, 
respectively (see Background section for definition). The parameter range for the family is $p \in (0,\infty)$, lower-tail dependence is $2^{-1/p}$ and upper-tail
dependence is $0$.\label{figped_C}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=\textwidth]{./Results/PedagogSuppMat_BB1.pdf}
\caption[Example BB1 copulas]{Log-transformed pdfs for example BB1 copulas. K is Kendall correlation; $p_1$ and $p_2$
denote the two parameters of the family (it is a 
two-parameter family); and LT and UT are our measures of lower- and upper-tail dependence, 
respectively (see Background section for definition). The parameter ranges for 
the family are $p_1 \in (0,\infty)$ and $p_2 \in [1,\infty)$, 
lower-tail dependence is $2^{-1/(p_1 p_2)}$ and upper-tail
dependence is $2-2^{1/p_2}$.\label{figped_BB1}}
\end{center}
\end{figure}

@@

We addressed Q1 first via a model selection procedure in which several
families of copulas were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). One of the fitted copulas was the N
copula, making possible comparisons of the degree to which the N copula versus
other copula families were good descriptions of data.

For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, we produced normalized ranks 
$u_i$ and $v_i$ as in the Introduction. 
Second, we tested the independence of the $u_i$ and $v_i$
using the statistic $T=\sqrt{(9N(N-1))/(4N+10)}|\tau|$, where $\tau$
is Kendall's tau for the data \cite{Genest2007}. We used the 
implementation of this test given in the `BiCopIndTest` in the `VineCopula` package
in the R programming language. 
We tested for independence because our model selection 
algorithms were ineffective if data could not 
be distinguished from independent data 
since many of the copula families we considered include the independent copula. 
If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit 16 bivariate copula families (see below for 
information on the families used) to the 
normalized ranks via maximum likelihood. The approach of fitting
copula families to normalized ranks was studied in depth
and recommended by @Genest1995 and @Shih1995. Their estimator of copula
parameters, which we use, was shown to be consistent, asymptotically normal,
and fully efficient at indepencence [@Genest1995]. @Genest2007
recommend carrying out inferences of dependence structures (which was our
goal here) using ranks.
We used the fitting implementation given in `BiCopEst` in `VineCopula`.
Fourth, we obtained AIC and BIC values and accompanying model weights 
$\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection] for each fitted copula. 
`BiCopEst` also 
provided lower- and 
upper-tail dependence of the best-fitting member of each family. 
$\text{AIC}_{\text{w}}$ values were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
